Defining properties regulating pioneer-factor binding and function

### 2022-12-08 manuscript/manuscript.qmd
update MS results

To characterize the extent of chromatin opening by Zld and Grh, we integrated our ChIP-seq and ATAC-seq data and identified 3 classes of binding sites (Fig. 1a,b). Class I sites are accessible in WT cells and remain accessible when bound by ectopically-expressed Zld or Grh. Class II sites are inaccessible in WT cells, however binding of Zld or Grh does not lead to chromatin opening. Class III sites are inaccessible in WT cells and Zld or Grh binding results in chromatin opening. The majority of ChIP-seq peaks for Zld and Grh were class I sites, with a subset of binding sites in closed chromatin, and a small subset undergoing chromatin opening (Fig. 1c,f). While class I binding sites occured mostly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig. 1d,g). Genome-wide analysis of signa <!--#  -->l intensities at all class II and III sites confirmed that these sites lack ATAC signal in WT cells and that class III sites undergo robust chromatin opening upon Zld or Grh induction (Fig. 1e,h). Together, these data show that when expressed exogenously at physiolgical levels, Zld and Grh can bind and open closed chromatin at a subset of binding sites.

@@

Having established that ectopically expressed Zld and Grh bind and open closed chromatin, we sought to determine if binding of these factors also led to transcriptional activation. To identify changes in gene expression upon Zld or Grh induction, we performed RNA-seq and compared induced cells to WT cells lacking Zld or Grh expression. Differential expression analysis using DESeq2 [@love2014] identified XXX differentially expressed genes for Zld and XXX for Grh (Extended Data Fig. XX).

@@

-   Twi binds closed chromatin and opens at some sites

@@

-   Twi also promotes transcriptional activation.

@@

>>>>

To characterize the extent of chromatin opening by Zld and Grh, we integrated our ChIP-seq and ATAC-seq data and identified three classes of binding sites (Fig. 1a,b). Class I sites are accessible in WT cells and remain accessible when bound by ectopically-expressed Zld or Grh. Class II sites are inaccessible in WT cells, however binding of Zld or Grh does not lead to chromatin opening. Class III sites are inaccessible in WT cells and Zld or Grh binding results in chromatin opening. The majority of ChIP-seq peaks for Zld and Grh were class I sites, with a subset of binding sites in closed chromatin, and a small subset undergoing chromatin opening (Fig. 1c,f). While class I binding sites occured mostly at promoters, the majority of class II and III sites were at promoter-distal sites (Fig. 1d,g). Genome-wide analysis of signa <!--#  -->l intensities at all class II and III sites confirmed that these sites lack ATAC signal in WT cells and that class III sites undergo robust chromatin opening upon Zld or Grh induction (Fig. 1e,h). Together, these data show that when expressed exogenously at physiolgical levels, Zld and Grh can bind and open closed chromatin at a subset of binding sites.

@@

Having established that ectopically expressed Zld and Grh bind and open closed chromatin, we sought to determine if binding of these factors also led to transcriptional activation. To identify changes in gene expression upon Zld or Grh induction, we performed RNA-seq and compared induced cells to WT cells lacking Zld or Grh expression. Differential expression analysis using DESeq2 [@love2014] identified XXX differentially expressed genes for Zld and XXX for Grh (Extended Data Fig. XX). Upregulated genes were significantly enriched for Zld or Grh binding, while downregulated genes are not (Extended Data Fig. XX <!--# include statistics  -->), suggesting that binding of these factors mainly leads to transcriptional activation and not repression. <!--# Consider mentioning Grh as a repressor here and giving examples of genes where Grh may be repressing. Discuss if Grh-mediated transcriptional repression coincides with chromatin opening, chromatin closing, or neither --> Gene ontology (GO) analysis indicated that the genes activated by each factor are consistent with the endogenous functions of Zld and Grh (Extended Data Fig. XX). <!--# Add more detailed information on GO analysis here -->

To better understand the relationship between chromatin opening and transcriptional activation, we directly compared our RNA-seq and ATAC-seq data. Overall, chromatin opening was correlated with transcriptional activation (Fig. 1i-l). sites with significantly increased chromatin accessibility, but not decreased or unchanged sites, had larger RNA-seq fold change values (Extended Data Fig. XX <!--# add statistics -->). When we classified our three classes of ChIP-seq based on changes in gene expression, we found that class III sites were most likely to be proximal to upregulated genes (Fig. 1j,l). This was particularly evident at class III promoters, which were most likely to correspond to upregulated genes. Together, we conclude that while binding and chromatin opening can happen in the absence of transcriptional changes, chromatin opening is correlated with transcriptional activation.

@@

-   Twi binds closed chromatin and opens at some sites.

-   Mention analysis of published Abd-A and Ubx cell lines.

@@

-   Twi also promotes transcriptional activation

@@

-   Mention differences between Twi vs. Zld and Grh

Despite the differences between Zld and Grh during Drosophila development, both proteins behaved similarly when expressed in our exogenous S2 cell system. To provide additional context to our findings, we performed additional experiments with the transcription factor Twist (Twi). Although Twi is also essential for regulating Drosophila embryonic development, Twi has a number if distnict properties from Zld and Grh, making it an interesting point of comparison. Twi is a master regulator of mesodermal cell fate (REFS). Twi has many binding sites that are dynamic during early embryonic development [@Wilczynski2010a], in contrast to Zld and Grh. Although Twi was hypothesized to be a pioneer factor [@Sandmann2007], pioneering activity by Twi had not previously been tested. In the early Drosophila embryo, mutation of Zld binding sites leads to a loss of Twi binding to the *cactus* enhancer, suggesting that pioneering by Zld may be required for Twi binding in the embryo [@Yáñez-Cuna2012]. To test if Twi is a pioneer factor, we generated additional stable cell lines allowing inducible expression of Twi at approximately physiological levels (Extended Data Fig. XX). We characterized Twi binding and activity using ChIP-seq, ATAC-seq, and RNA-seq. Comparison of Twi ChIP-seq and ATAC-seq data allowed for the identification of the same classes of binding sites that were found for Zld and Grh (Fig 2a,b). Similar to Zld and Grh, class I binding sites were enriched at promoters, while class II and III binding sites were mostly promoter-distal (Fig 2c). Twi bound extensively to closed chromatin, with XX% of binding sites lacking detectible chromatin accessibility (Fig. 2b). At a small subset of Twi binding sites, binding to closed chromatin resulted in chromatin opening (Fig 2a-c). Analysis of RNA-seq data revealed that Twi expression resulted in transcriptional activation of hundreds of genes (Extended Data Fig. XX). Upregulated genes were significantly enriched for Twi binding relative to unchanged genes ((Extended Data Fig. XX).
